---
title: "Analysis of bikeshare data"

author: "Joe Skufca"
date: "19 Sep 2020"
output: html_notebook
---

The goal of this analysis is to provide some prelimiary analysis on the bike share data for August 2020, primarily through visualization.

## Prepare workspace:

#### Load packages

We will work primarily within the context of the tidyverse set of packages, with a few additional packages supporting exploratory analysis.  I have included the `lubridate` package, as we will want to do some work with dates.

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(skimr)
library(summarytools)
library(lubridate)
```

I will also set a default theme for my ggplots.

```{r}
theme_set(theme_minimal())
```


## Data

I am **not** using the file that (a) started on moodle for which we,  (b) computed a couple of extra variables (for duration, hour of day, and day of week), and (c) saved as .xlsx file.  

The orginal source of the data was the csv file 

https://s3.amazonaws.com/capitalbikeshare-data/202008-capitalbikeshare-tripdata.zip

so I will use that original source data so that my research is more *reproducible*.

>Note 1: My variable names might be different than yours.  Be careful to think through what I am doing and use it as a template.




### Read the data

I read the original .csv file and add in the variables (mutate) that we did manually when we played with the creation of the data table.   I will call these new variables: `duration_min`, `hour_of_day`.  Also I will add `day of week`, but it won't match the format from excel.

```{r}
dfa= read_csv("202008-capitalbikeshare-tripdata.csv")%>% 
  clean_names() %>%
  mutate(duration=as.numeric((ended_at-started_at)/60),
         hour_of_day=hour(started_at),
         day_of_week=wday(started_at,label = T))
  
```


We summarize that dataframe to identify data types, missing data, et cetera.

```{r}
skim(dfa)
```

#### A histogram - like we did in class

As we discussed in class, we might be interested in looking at histogram of durations, broken out for each day of the week.  Here I filter to positive duration shorter than 100 minutes.

```{r}
dfa %>% filter(duration<100,duration>0) %>% 
  ggplot(aes(x=duration)) + geom_histogram(bins = 300)+facet_wrap(~day_of_week)
```


### Cleaning

Before we do further analysis, we recognize that if the duration is negative, some piece of the time data is corrupt.  Although it sill reduce the size of our dataset, it is very much large enough to permit continued analysis even after removing those rows.

Limit to durations that are positive.

```{r}
dfb=dfa %>% filter(duration>0)
```

### Some visual exploration.

Can a scatterplot be revealing?  Many possibilities, but I have decided to focus on "short" trips (duration less than an hour) for a particular day (the 3rd day of the month).  

Does the trip duration have a pattern when plotted vs. starting time?

```{r}
dfb %>% filter(mday(started_at)==3,duration<60) %>% ggplot(aes(x=started_at,y=duration))+
  geom_point(alpha=.1)+
  ggtitle("Trip duration vs start time (August 3)")
```

### Riders vs time

Let's try to construct a dataframe that could track number of riders (at any instant in time), where we will assume (not reasonably) that there were no riders at 00:00.


I will develop (and test) the logic on a small set of data.  
A small dataset works fast and is easy to read.  
Once we build the logic for the small set, it is easy to generalize.


The `slice_sample` command below takes a random sample of 100 rows.
So ... my experimental dataframe, `dfe`,  limits to the 3rd day of the month, and only 100 rows.


```{r}
dfe=dfb %>% filter(mday(started_at)==3) %>% slice_sample(n=100)
```


#### Designing an algorithm

Now lets build a process.   I will proceed step by step (copying and pasting to add one new step to my process until I get what I want).



I think we only need start and end times.

```{r}
dfe %>% select(start=started_at,end=ended_at)
```
Create a long dataset

```{r}
dfe %>% select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end) 
```
Pick good names, then order by time.

```{r}
dfe %>% select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t)
```

I want to build a counter to track riders, where each "start" increases the count, and each end "decrease" the count.

Additionally, I will want my to `arrange` my rows using my time `t` column.

```{r}
dfe %>% select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) 
```

Now lets cummulative sum that column.

```{r}
dfe %>% select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) %>%
  mutate(riders=cumsum(increment))
```
I think that is what I need.  I will visualize (using a step plot) to see if it makes sense.   I simply ... continue the pipe.

```{r}
dfe %>% select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) %>%
  mutate(riders=cumsum(increment)) %>% 
  ggplot(aes(t,riders)) + geom_step()
```

I note the quirk that one ride started on day three and didn't get returned until the fifth of August, but ... I am convinced by logic works.

#### Analysis without sampling:

Now ... let's examine all the data for August 3rd.  I simply need to drop in the dataframe for August 3rd *without* doing the steps that select a random sample, but *otherwise using the exact commands developed above.*

For the visualization, I will limit to only 3 August, which is a bit tricky.


```{r}
dfb %>% filter(mday(started_at)==2) %>%
  select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) %>%
  mutate(riders=cumsum(increment)) %>% 
  ggplot(aes(t,riders)) + geom_step() +
  scale_x_datetime(limits=as_datetime(   c("2020-08-02","2020-08-3")))
```

We could look at the whole dataset.  All we have to do is *not* do that filtering step ... and let the axis cover the whole range.


```{r}
dfb %>% 
  select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) %>%
  mutate(riders=cumsum(increment)) %>% 
  ggplot(aes(t,riders)) + geom_step() 
```

I think this is interesting, but I think I can improve using faceting, to separate out for each day of the month.   I will facet by 7 columns so that days of the week are aligned.  For these facets, I use the option "free_x" as each facet is covering a different time range.

It also turns out that the datset includes information from 1,2,3,4 September, which screws things up a bit, so ... I filter to only the month of August.


```{r}
dfb %>% 
  filter(month(started_at)==8,month(ended_at)==8,mday(started_at)<=14) %>%
  select(start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) %>%
  mutate(riders=cumsum(increment)) %>% 
  ggplot(aes(t,riders)) + geom_step() +
  facet_wrap(~mday(t),scales = "free_x",ncol = 7)+
  theme(axis.text.x = element_blank())
```

#### Separating electric and docked bikes


To consider by type of bike, we will need "cumsum" to be working separately on "electric bikes" and "docked bikes".  Which means:

* We have to include that variable in the dataset, and
* We will need to group on that variable

For this visualization, although I do the compuation for the whole month,
I will show only the first 7 days of August.


```{r}
dfb %>% 
  filter(month(started_at)==8,month(ended_at)==8) %>%
 select(rideable_type,start=started_at,end=ended_at) %>%
  pivot_longer(start:end, names_to="type",values_to="t") %>% arrange(t) %>%
  mutate(increment=case_when(
   type=="start"~1,
   type=="end" ~ -1
  )) %>%
  group_by(rideable_type) %>%
  mutate(riders=cumsum(increment)) %>% filter(mday(t)<=7) %>%
    ggplot(aes(t,riders,color=rideable_type)) + geom_step() +
  facet_wrap(~mday(t),scales = "free_x",ncol = 7)
```

### Box and whisker plot for trip duration based on "hour".


Because I said we could.  Note -  I specified "hour of day" as a grouping variable.


```{r}
dfb %>% filter(duration<60) %>% ggplot(aes(x=hour_of_day,y=duration,group=hour_of_day)) +
  geom_boxplot()
```

