---
title: "Covid and population data"
output: html_notebook
---

Quick try to get data and merge.

```{r}
library(tidyverse)
library(coronavirus)
library(wbstats)

```
We will make use of the **DT** package for interactive display of tabular output in the Appendix.

```{r load-packages, warning=FALSE, message=FALSE}
library(coronavirus) # devtools::install_github("RamiKrispin/coronavirus")
library(tidyverse)
library(lubridate)
library(glue)
library(scales)
library(ggrepel)
library(DT)
```

## Data prep

The data frame called `coronavirus` in the coronavirus package provides a daily summary of the Coronavirus (COVID-19) cases by country. Each row in the data frame represents a country (or, where relevant, state/province). A full list of the countries in the data frame is provided in the [Appendix]. Note that the data provided in this package provides daily number of deaths, confirmed cases, and recovered cases. For this report, we will focus on the deaths. 

We will start by making our selection for the countries we want to explore.

```{r select-countries}
countries <- c(
  "China",
  "France",
  "United Kingdom",
  "US",
  "Turkey"
)
```

In the following code chunk we filter the data frame for deaths in the countries we specified above and calculate cumulative number of deaths. We will only visualise data since 10th confirmed death. 

```{r prep-country-data}
country_data <- coronavirus %>%
  # filter for deaths in countries of interest
  filter(
    type == "death",
    country %in% countries
  ) %>%
  # fix county labels for pretty plotting
  mutate(
    country = case_when(
      country == "United Kingdom" ~ "UK",
      TRUE ~ country
    )
  ) %>%
  # calculate number of total cases for each country and date
  group_by(country, date) %>%
  summarise(tot_cases = sum(cases)) %>%
  # arrange by date in ascending order
  arrange(date) %>%
  # record daily cumulative cases as cumulative_cases
  mutate(cumulative_cases = cumsum(tot_cases)) %>%
  # only use days since the 10th confirmed death
  filter(cumulative_cases > 9) %>%
  # record days elapsed, end date, and end label
  mutate(
    days_elapsed = as.numeric(date - min(date)),
    end_date     = if_else(date == max(date), TRUE, FALSE),
    end_label    = if_else(end_date, country, NULL)
  ) %>%
  # ungroup
  ungroup()
```

We also need to take a note of the "as of date" for the data so that we can properly label our visualisation.

```{r record-as-of-date}
as_of_date <- country_data %>% 
  summarise(max(date)) %>% 
  pull()
as_of_date_formatted <- glue("{wday(as_of_date, label = TRUE)}, {month(as_of_date, label = TRUE)} {day(as_of_date)}, {year(as_of_date)}")
```

These data are as of `r as_of_date_formatted`.

## Visualisation

The following visualisation shows the number of cumulative cases vs. days elapsed since the 10th confirmed death in each country. The time span plotted for each country varies since some countries started seeing (and reporting) deaths from COVID-19 much later than others.

```{r visualise, warning=FALSE}
ggplot(data = country_data,
       mapping = aes(x = days_elapsed, 
                     y = cumulative_cases, 
                     color = country, 
                     label = end_label)) +
  # represent cumulative cases with lines
  geom_line(size = 0.7, alpha = 0.8) +
  # add points to line endings
  geom_point(data = country_data %>% filter(end_date)) +
  # add country labels, nudged above the lines
  geom_label_repel(nudge_y = 1, direction = "y", hjust = 1) + 
  # turn off legend
  guides(color = FALSE) +
  # use pretty colors
  scale_color_viridis_d() +
  # better formatting for y-axis
  scale_y_continuous(labels = label_comma()) +
  # use minimal theme
  theme_minimal() +
  # customize labels
  labs(
    x = "Days since 10th confirmed death",
    y = "Cumulative number of deaths",
    title = "Cumulative deaths from COVID-19, selected countries",
    subtitle = glue("Data as of", as_of_date_formatted, .sep = " "),
    caption = "Source: github.com/RamiKrispin/coronavirus"
  )
```

## Appendix

A list of countries in the `coronavirus` data frame is provided below.

```{r list-countries, echo=FALSE}
coronavirus %>%
  select(country) %>%
  arrange(country) %>%
  distinct() %>%
  datatable()
```



```{r}
dfa=coronavirus

country_df=dfa
```

```{r}
dfb=wb_data("SP.POP.TOTL") %>% filter(date==2019)
```

